### 2022-02-15 introduction/introduction.tex
more

The initial study of this formalism and the rules to reason about statements in
it was very prolific. In 1930 G\"odel \cite{kg-30-completeness-theorem} showed
that there is a system of inference rules (calculus) such that for any valid
statement a proof of validity can be constructed in this system. Around the same
time, the first computers were leaving the imagination of the engineers and
began to automate many complicated processes.  Naturally, the question as to
whether a computer can decide if a first-order statement is valid (i.e., a
theorem) arose. Church \cite{ac-36-fol-undecidable} and Turing
\cite{tm-37-undecidable} answered this question negatively in 1936.

@@

Even though the successes of proving Robbins conjecture showed that it is
possible, automatically proving hard mathematical problems is still out of reach
for theorem provers. However, computers can still be of help to construct formal
proofs.

Proof assistants (or interactive theorem provers) are programs which allow users
to describe a theory formally and to write proofs of the statements about this theory
using inference system rules of the proof assistant. These
programs can then check if the given proof is correct. As their core is usually
small and well-tested, proofs that pass the checking phase are considered
trustworthy.

@@

This approach proved useful: around 69\% of the simple proof goals can be
discharged automatically using Sledgehammer \cite{kb-13-mash}. However, there is
clearly some untapped potential in the approach as it communicates with
automatic theorem provers through a translation. To bridge this translation gap,
attempts were made to use higher-order automatic theorem provers instead of
first-order ones \cite{ns-13-leo2sh}. This proved unfruitful as state-of-the-art
higher-order of the time theorem provers were primarily designed for small and tricky
problems. On the other hand, problems coming from hammers consist of mostly first-order formulas, are rather large,
and mostly require simple higher-order proof steps.

Thus, from a standpoint of proof assistant ideal automatic theorem prover fulfills
the following wishlist:
\begin{itemize}
  \item Performs as best first-order provers
  \item Supports higher-order logic
  \item Works well on large problems
\end{itemize}
This brings us to the topic of this thesis: \emph{How can we develop a theorem provers that fulfills this wishlist?}

>>>>

\paragraph{Automatic Theorem Proving}The initial study of first-order logic and the rules to reason about statements in
it was very prolific. In 1930 G\"odel \cite{kg-30-completeness-theorem} showed
that there is a system of inference rules (calculus) such that for any valid
statement a proof of validity can be constructed in this system. Around the same
time, the first computers were leaving the imagination of the engineers and
began automating many complicated processes.  Naturally, the question as to
whether a computer can decide if a first-order statement is valid (i.e., a
theorem) arose. Church \cite{ac-36-fol-undecidable} and Turing
\cite{tm-37-undecidable} answered this question negatively in 1936.

@@

Even though the successes of proving Robbins conjecture showed that it is
possible, automatically proving hard mathematical problems is still out of reach
for theorem provers. However, there are other ways in which humans and computers can work together to construct formal
proofs.

\paragraph{Proof Assistants and Hammers} Proof assistants (or interactive
theorem provers) are programs which allow users to describe a theory formally
and to write proofs of the statements about this theory using inference 
rules of the proof assistant. These programs can then check if the given proof
is correct. As their core is usually small and well-tested, proofs that pass the
checking phase are considered trustworthy.

@@

This approach proved useful: around 69\% of the simple proof goals can be
discharged automatically using Sledgehammer \cite{kb-13-mash}. However, there is
clearly some untapped potential in the approach as it communicates with
automatic theorem provers through a translation. To bridge this translation gap,
attempts were made to use higher-order automatic theorem provers instead of
first-order ones \cite{ns-13-leo2sh}. This proved unfruitful as state-of-the-art
higher-order theorem provers of the time were primarily designed for small and
tricky problems. On the other hand, problems coming from hammers consist of
mostly first-order formulas, are rather large, and mostly require simple
higher-order proof steps. Thus, from a standpoint of proof assistant ideal
automatic theorem prover fulfills the following wishlist:
\begin{enumerate}
  \item\label{it:fo} Performs as best first-order provers on first-order problems
  \item\label{it:ho} Supports higher-order logic and scales with the amount of higher-order axioms
  \item\label{it:size} Scales well with the size of the problem
\end{enumerate}
This brings us to the topic of this thesis: \emph{How can we develop a theorem provers that fulfills this wishlist?}

\paragraph{Our Approach} To improve the automatic reasoning for higher-order
logic we decided not to start from a blank slate and build a new theorem prover.
Rather, we start from a position of strength --- from state-of-the-art
first-order superposition theorem prover E \cite{scv-19-e23} --- and extend this
prover to higher-order logic. In this way, the points \ref{it:fo} and
\ref{it:size} from the above wishlist  is fulfilled for free, as E is very
efficient and deals well with large problems.

We do not extend E to higher-order logic in one atomic step. Instead, we extend
it by adding support for three distinguished features of higher-order logic,
which gradually increase its support for higher-order logic. With the layered
approach we make sure that after every extension to higher-order logic, E's
efficiency on first-order problems was not affect. In other words, while trying
to fulfill the point \ref{it:ho} from the wishlist we make sure that points
\ref{it:fo} and \ref{it:size} are not affected.

We have split the road to full higher-order logic in three stages. A
distinguished feature of higher-order logic is that it not only supports
quantification over objects, but also over functions that manipulate these
objects. Thus, higher-order logic allows us to make a statement ``all functions
with arguments $0$ and $1$ result in $2$" ($\forall x. x (\cst{0}, \cst{1}) \eq
\cst{2}$). Replacing $x$ with addition function ($\cst{+}$) is enough to
disprove this statement. The first stage is to extend E to support
quantification over functions that are already present in the theory.

Higher-order logic also supports expressing functions that are not explicitly
stated in the problem. For example, suppose we want to prove that there is a
function such that returns 4 when given 2 and 5 when given 3 ($\exists x.
x(\cst{2}) \eq \cst{4} \lland x (\cst{3}) \eq \cst{5} $). An example of such
function is the function that adds 2 to its argument. In higher-order logic such
unnamed function can be represented using $\lambda$ abstractions as $\lambda
x.\, x \, \cst{+} \, 2$. Replacing $x$ with this term suffices to prove the
original statement. The second stage is to support automatically synthesizing
such functions.

In higher-order logic we can make statements about statements. For example, we
can naturally represent statements such as ``whatever Jasmin says is true'' as
$\forall x. \, \cst{says}(\cst{jasmin}, x) \imp x$. The last stage is to support this
feature of first-order logic.
